---
title: "Accessing elevation data in R with the elevatr package"
author: "Jeffrey W. Hollister"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to elevatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE, echo=FALSE}
################################################################################
#Load packages
################################################################################
library("sp")
library("raster")
library("rgdal")
```

# Introduction to `elevatr`

Elevation data is used for a wide array of applications, including, for example, visualization, hydrology, and ecological modelling.  Gaining access to this data in R has not had a single interface and was made available through functions across many packages or required local access to the data.  This is no longer required as a variety of APIs now exist that provide programmatic access to elevation data.  The `elevatr` package was written to standarize access to elevation data from web APIs.  This introductory vignette provides details on how to use `elevatr` to access elevation data and provides a bit of detail on the source data it accesses.

There are currently four endpoints that `elevatr` accesses.  For point elevation data it uses USGS Elevation Point Query Service and Mapzen Elevation Service and to access raster elevation data (e.g. a DEM) it uses Mapzen Terrain Service, and the Amazon Web Services Terrain Tiles.

# Get Point Elevation Data

Point elevation is accesses from `get_elev_point()`.  This function takes either a data.frame of XY locations or a SpatialPoints/SpatialPointsDataFrame as input and then fetches the reported elevation for that location.  As mentioned there are two services that provide this information.  Details and use for those are provided below.

## Mapzen Elevation Service 

The [Mapzen Elevation Service](https://mapzen.com/documentation/elevation/elevation-service/) provides point elevations that are currently derived from the  [SRTM](http://www2.jpl.nasa.gov/srtm/), [GMTED](http://topotools.cr.usgs.gov/gmted_viewer/) and [GEBCO](http://www.gebco.net/data_and_products/gridded_bathymetry_data/) digital elevation models.  The web API provides a variety of input and output sources, but `elevatr` only uses the shape list and returns only the elevation.  The [range parameter](https://mapzen.com/documentation/elevation/elevation-service/#get-height-and-distance-with-the-range-parameter) may be implemented at some point in the future.

### API Keys

A Mapzen API Key is not required to use this service, but the rate limit is 1000 requests per day, 6 per minute, and 1 per second which is impractical for typical uses, thus it is suggested that you get and use an API Key.  With the API Key the rate limit is 2 queries per second and 20000 queries per day.

Get an API Key from <https://mapzen.com/developers>.  Since `elevatr` will search the environment for the appropriate key , it is suggested that you store this in your `.Renviron` file.  To do this:

```{r api_key, eval=FALSE}
cat("mapzen_key=mapzen-XXXXXXX\n",
    file=file.path(normalizePath("~/"), ".Renviron"),
    append=TRUE)
```

Once this is set for a given machine there is no need to include the API key in the functions themselves as the default is to use the `mapzen_key` environment variable.

### Using `get_elev_point()` to Access The Mapzen Elevation Service

Usage of `get_elev_point()` requires an input SpatialPoints, SpatialPointsDataFrame, or a two-column data frame with column one containing the x (e.g. longitude) coordinates and the second column containg the y coordinates (e.g. latitude).  The source data is global and also includes estimates of depth for oceans and major lakes

Example usage of each is included below.  For these examples, we can create a dataset to use.

```{r example_dataframe}
# Create an example data.frame
set.seed(65.7)
examp_df <- data.frame(x = runif(10, min = -73, max = -71), 
                       y = runif(10, min = 41 , max = 45))
prj_dd <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

# Create and example data.frame with additional columns
cats <- data.frame(category = c("H", "H", "L", "L", "L", 
                                "M", "H", "L", "M", "M"))

examp_df2 <- data.frame(examp_df, cats)

# Create an example SpatialPoints
examp_sp <- SpatialPoints(examp_df, proj4string = CRS(prj_dd))

# Create and example SpatialPointsDataFrame

examp_spdf <- SpatialPointsDataFrame(examp_sp, proj4string = CRS(prj_dd), data = cats )
```

If a data frame is used it may have additional columns beyond the first two, which must contain the coordinates.  The additional columns, along with the returned elevation, will be part of the output SpatialPointsDataFrame.  Similarlily, an elevation column is added to the data slot of a SpatialPointsDataFrame.

```{r examples_df1}
# Example using data.frame with longitude and latitude
df_elev <- get_elev_point(examp_df, prj = prj_dd, src = "mapzen")

# Compare
examp_df
df_elev
```

```{r examples_df2}
# Example using data.frame with longitud, latitude and an additional column
df2_elev <- get_elev_point(examp_df2, prj = prj_dd, src = "mapzen")

# Compare
examp_df2
df2_elev
```

```{r sleep1, echo=F}
Sys.sleep(1)
```

The process is the same for a SpatialPoints and a SpatialPointsDataFrame object.

```{r examples_sp, eval=T}
# Example using SpatialPoints
# prj is taken from the SpatialPoints object
# api_key is taken from environment variable mapzen_key
sp_elev <- get_elev_point(examp_sp)

# Compare
examp_sp
sp_elev
```

```{r examples_sp2, eval=T}
# Example using SpatialPoints
# prj is taken from the SpatialPoints object
# api_key is taken from environment variable mapzen_key

spdf_elev <- get_elev_point(examp_spdf)

# Compare
examp_spdf
spdf_elev
```

## USGS Elevation Point Query Service 

The [USGS Elevation Point Query Service](http://ned.usgs.gov/epqs/) is also accessible from `elevatr`.  This service is included as it provides elevation from a single source (as opposed to the [muliple sources from Mapzen](https://mapzen.com/documentation/elevation/elevation-service/#data-sources-and-known-issues)) and provides higher precision elevation.  It is only available for the United States (including Alaska and Hawaii).  Points that fall within the United States but are not on land return a value of zero.  Points outside the United States boundaries return a value of -1000000.

### Using `get_elev_point()` to Access The USGS Elevation Point Query Service

The USGS Elevation Point Query Service returns a single point at a time.  The implemntation in `get_elev_point()` will loop through each point, thus can be slow for large number of requests.  In those instances, the Mapzen source would be preferred.

Accessing data from this service is done by setting the `src` to `"epqs"`.  No API key is required and there are no rate limits.

```{r}
df_elev_epqs <- get_elev_point(examp_df, prj = prj_dd, src = "epqs")
df_elev_epqs
df2_elev_epqs <- get_elev_point(examp_df2, prj = prj_dd, src = "epqs")
df2_elev_epqs
sp_elev_epqs <- get_elev_point(examp_sp, src = "epqs")
sp_elev_epqs
spdf_elev_epqs <- get_elev_point(examp_spdf, src = "epqs")
spdf_elev_epqs
```

# Get Raster Elevation Data

## Mapzen Terrain Tile Service

<https://mapzen.com/documentation/terrain-tiles/>

### API Keys

## Terrain Tiles on Amazon Web Services (AWS)

<https://aws.amazon.com/public-datasets/terrain/>

<!--
Begining of vignette for elevatr package

```{r, eval=F}
library(sp)
library(elevatr)
library(raster)
dim_df <- data.frame(lat=NA,z=NA,x_res=NA,y_res=NA)
ct <- 1
for(i in seq(0,60,10)){
  xdf <- SpatialPoints(coordinates(data.frame(0,i)),CRS("+proj=longlat"))
  xdf_alb <- spTransform(xdf,CRS("+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m +nadgrids=@null +wktext  +no_defs"))
  for(j in 0:14){
    x <- get_elev_raster(xdf_alb,z=j)
    dim_df[ct,]<-data.frame(i,j,res(x)[1],res(x)[2])
    ct <- ct + 1
  }
}
dim_df
```
--!>